Analysis of Programming Students' Prompts to Identify Interactions Patterns with Generative AI

### 2024-09-19 CSEDU2025/Paper.tex
update

Regardless of teachers' preferences or beliefs, preliminary surveys
performed by \cite{Dickey24} revealed that at least 54.5\% of students are
already using GenAI for homework. This highlights the need to understand how
students are interacting with these tools and how they can be used to enhance
learning. Furthermore, a review by \cite{Lo24} emphasized the necessity for
extended studies and objective measures to gather more robust evidence on the
use of GenAI tools in education.

@@

\subsection{Critical Thinking}

Critical thinking is a fundamental skill for students to develop, as it enables
them to analyze, evaluate, and synthesize information. This skill is essential
for problem-solving and decision-making, making it a crucial aspect of
educational success, specifically in programming courses. However, the use of AI
tools like ChatGPT has raised concerns about the potential impact on students'
critical thinking abilities \cite{Murillo23} \cite{cai23} \cite{chan23}.

In the other hand, the study of \cite{zhang24} suggested that incorporating
ChatGPT could stimulate students' critical thinking, probably because they
had to assess the correctness of ChatGPT's answers.

Regarding self-regulation, \cite{cai23} noted that ChatGPT allowed students to
regulate their learning pace.

\cite{wu24} found that ChatGPT-supported learning not only promoted students'
self-regulation but also increased their self-efficacy.

@@

\begin{table*}[htbp]
  \caption{Categories of Analysis}
      \begin{center}
      \begin{tabular}{|p{4cm}|p{6cm}|p{3cm}|}
      \hline
      \textbf{Category} & \textbf{Description} & \textbf{Critical Thinking} \\
      \hline
      Debugging Help & Prompts that seek help to identify, fix errors, or understand the provided code snippet. & \multirow{7}{3cm}{Tends to promote.} \\
      Conceptual Questions & Prompts that are more about understanding concepts or algorithms rather than specific code. & \\
      Student Correction & Prompts where the student corrects the bot. & \\
      \hline
      Code Snippet & Prompts that ask for a specific part of the code, like a function or a segment. & \multirow{6}{3cm}{Tends not to promote.} \\
      Complete Solution & Prompts that request an entire solution or a complete code snippet. & \\
      Multiple Question & Prompts where the user wants to solve a multiple-question exercise. & \\
      \hline
      Language Change & Prompts that request a change of idiom to convey a message more effectively. & \multirow{4}{3cm}{Neutral.} \\
      Uncategorized & Prompts that do not fit into any of the above categories. & \\
      \hline
      \end{tabular}
      \label{tab:categories}
      \end{center}
  \end{table*}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.55]{img/figure1.png}
    \caption{Classification of messages.}
    \label{fig:graph1}
\end{figure}

Approximately 64\% of the messages pertain to categories associated with
critical thinking, corroborating with the findings of
\cite{Ghimire24}. In contrast, around 28\% of the messages
indicate a preference among students for ready-made answers.

>>>>

Regardless of teachers' preferences or beliefs, a preliminary surveys performed
by \cite{Dickey24} revealed that at least 54.5\% of students are already using
GenAI for homework. This highlights the need to understand how students are
interacting with these tools and how they can be used to enhance learning.
Furthermore, a review by \cite{Lo24} emphasized the necessity for extended
studies and objective measures to gather more robust evidence on the use of
GenAI tools in education.

@@

% Critical thinking is a fundamental skill for students to develop, as it enables
% them to analyze, evaluate, and synthesize information. This skill is essential
% for problem-solving and decision-making, making it a crucial aspect of
% educational success, specifically in programming courses. However, the use of AI
% tools like ChatGPT has raised concerns about the potential impact on students'
% critical thinking abilities \cite{Murillo23} \cite{cai23} \cite{chan23}.

% In the other hand, the study of \cite{zhang24} suggested that incorporating
% ChatGPT could stimulate students' critical thinking, probably because they
% had to assess the correctness of ChatGPT's answers.

% Regarding self-regulation, \cite{cai23} noted that ChatGPT allowed students to
% regulate their learning pace.

% \cite{wu24} found that ChatGPT-supported learning not only promoted students'
% self-regulation but also increased their self-efficacy.

\begin{table*}[htbp]
  \caption{Categories of Analysis}
      \begin{center}
      \begin{tabular}{|p{4cm}|p{9cm}|}
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help & * Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question & * Prompts that are more about understanding concepts or algorithms rather than specific code. \\
      Student Correction & * Prompts where the student corrects the bot. \\
      \hline
      Code Snippet & * Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution & * Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question & * Prompts where the user wants to solve a multiple-question exercise. \\
      \hline
      Language Change & * Prompts that request a change of idiom to convey a message more effectively. \\
      Uncategorized & * Prompts that do not fit into any of the above categories. \\
      \hline
      \end{tabular}
      \label{tab:categories}
      \end{center}
  \end{table*}

\subsection{Metacognition}

The use of GenAI tools can help students learning by providing rapid assistance
and feedback. However, it is important to teach students how to use these tools
properly to promote learning. In this sense, it is important that students are
aware of the cognitive processes.

Metacognition is defined as the awareness about the cognitive processes.
\cite{flavell79} defined a model of cognitive monitoring in four classes of
phenomena that can act or interact with each other, which are: Metacognitive
Knowledge, Metacognitive Experience, Metacognitive Task, and Metacognitive
Strategies. Metacognitive knowledge refers beliefs about what factor or
variables affect the course and outcome of cognitive enterprise. It is
subdivided into three majors categories or variables: Person, Task, and
Strategy. Metacognitive experience refers to the feelings and emotions
associated with cognitive activities. Metacognitive task refers to the type of
task or problem that the individual is facing. Metacognitive strategies refer to
the methods or techniques that individuals use to control their cognitive
processes.

@@

@@

\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.6]{img/figure1.png}
    \caption{Classification of messages.}
    \label{fig:graph1}
\end{figure}

% Approximately 64\% of the messages pertain to categories associated with
% critical thinking, corroborating with the findings of
% \cite{Ghimire24}. In contrast, around 28\% of the messages
% indicate a preference among students for ready-made answers.

% Uncategorized

The Figure \ref{fig:graph1} shows that around 5\% of the prompts
written by students were not categorized. These messages range from
greetings to the Bot, such as a simple "\textit{Hola}," to
contextual statements like "\textit{Estoy repasando orientación a
objetos.}"

% Multiple Choice
\subsubsection*{Sequence of Multiple Choice}

Around 16.9\% of the messages were classified as Multiple Choice. These prompts
typically involve students sending one or more exercises to CharlieBot, asking
for the solution.

We identified a sequence of Multiple Choice patterns in some of the
conversations. From the point of view of Metacognitive Knowledge, a possible
interpretation of this complex phenomenon is that students may believe in their
inability to answer several questions. They may also assume that the questions
share a common theme, and solving them together may help consolidate knowledge.
Additionally, they might understand that obtaining the answers quickly is more
efficient.

Regarding Metacognitive Experiences, students may experience feelings such as
cognitive overload, a sense of urgency, and/or confidence in ready-made answers,
which can often lead to relief and a reduction of frustration.

Concerning the Metacognitive about the Task, one interpretation may be that the
student simply wants to complete the task quickly to alleviate the workload,
and/or obtain a correct answer without focusing on deeper learning at
this moment.

In relation to Metacognitive Strategies, students are likely focusing on saving
time and/or possibly lacking planning skills. They may also perceive that
solving the questions on their own requires too much cognitive effort, but this
pattern might reveal a lack of progress monitoring and understanding of
concepts. After obtaining the answers, the metacognitive evaluation process
could lead to feelings of satisfaction, but further reflection may indicate that
the strategy was not effective in supporting deep learning. Depending on this
evaluation, it may influence future decision-making to change the strategy.

% Conceptual Questions
\subsubsection*{Sequence of Conceptual Question}

Messages classified as conceptual questions represent around 32\% of all
messages. Regarding Metacognitive Knowledge, about these sequences of prompts
related with conceptual questions, one possible interpretation may be that the
students are aware they have not fully mastered a concept and are trying to
understand it, demonstrating a good level of self-awareness. In this sense, the
student realizes that the problem requires an understanding of concepts in order
to advance in their studies. In this way, the student applies an active learning
strategy. About Metacognitive Experience, several types of feelings may be
associated at this moment, such as actively reflecting on the answers,
uncertainty and/or confusion when trying to understand the concept, frustration
for not progressing, and satisfaction as their questions are answered in a
way that facilitates understanding. Regarding the type of metacognition related
to the Task, the student recognized the difficult and may be trying to achieve a
deeper understanding of the evolved concepts to later apply in practice. With
respect to Metacognitive Strategies, the student demonstrates a metacognitive
planning process when identifying their knowledge gaps. They may start with more
general questions to obtain an overview of the concept and then ask more
specific questions as they gain more clarity. Because the questions are in
sequence, the student probably can adequately monitor their progress.
